---
title: "BST 222 Homework 7"
author: "Jonathan Waring, Selena Huang, Laura Levin-Gleba, Wenze Tang"
date: "November 12, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Exponential Distribution

- The **exponential distribution** is the probability distribution that describes the time between events in a Poisson point process. 
- Exponential variables can be used to model situations where certain events occur with a constant probability per unit length, such as the distance between mutations on a DNA strand
- The probability density function (PDF) of an exponential distribution is given by $$f(x; \lambda) = \begin{cases}\lambda e^{-\lambda x} & x \geq 0 \\ 0 & x < 0 \end{cases}$$ where $\lambda > 0$ is the parameter of the distribution, often called the *rate parameter*

## Exponential PDF

```{r PDF}
library(ggplot2)
library(plyr)
library(latex2exp)
theme_set(theme_bw())

exponential <- function(lambda, x) {
  data.frame(x = x, value = lambda * exp(-lambda * x))
}

params <- expand.grid(lambda = c(0.5, 1, 1.5, 2))
pdfs <- mdply(params, exponential, x = seq(0,4, length=250))
ggplot(pdfs, aes(x, value, color=factor(lambda))) +
  geom_path() + 
  ylab(TeX("$f(x; \\lambda) = \\lambda e^{-\\lambda x} $")) + 
  labs(color=TeX("$\\lambda$"))
```

## Properties of $X \sim Exp(\lambda)$

- The **expected value** of  $X$ is given by $$E[X] = \frac{1}{\lambda}$$
- The **variance** of $X$ is given by $$Var[X] = \frac{1}{\lambda^2}$$
- The **median** of $X$ is given by $$m[X] = \frac{\ln (2)}{\lambda}$$


